An Object-Logic Explanation for the Inconsistency in G\"odel's Ontological Theory

### 2016-07-10 Papers/2016/KI/main.tex
Remove commented lines
Comment out appendix

%  Particularly interesting
% from the perspective of this paper is that many literature
% contributions unfortunately work with or refer to G\"odel's original
% definition of essence
% $$ todo $$
% This original version avoids the conjunct added by Scott expressing
% that essential properties of an individual should be possessed by the
% individual.

% We give here a list of examples:
% \begin{itemize}
% \item Anderson and Getting's 1996, p. 168 \cite[p.168]{AndersonGettings1968}
%   use essence without conjunct.
% \item  
% \item Look???, p. 514
% % \item Oppy 1996, p.226-227 \cite[p.226/227]{Oppy1996}, Oppy 2000, p. 364
% %   \cite[p.364]{Oppy200}, Oppy 2008, p. 1068 \cite[p.1068]{Oppy2008}:
% %   Oppy uses: ``A is an essence of x iff for every property B, x has B
% %   neces- sarily iff A entails B'' (this is from Anderson's
% %   emendation). Check is this leads to inconsistency as well. In this
% %   case we may write something like:

% %  Oppy presents an critical
% %   assessment in order to conclude that G\"odel's argument fails to
% %   convince. Oppy apparently does not see that he is assessing an
% %   inconsistent axiom system (and he in fact creates adaptations that
% %   suffer the same problems).
% % \item 
% \end{itemize}

@@

% Logic textbooks % \cite{ToDo:which}
% commonly utilize higher-order logic in an informal/semi-formal way as
% a meta-language to introduce the syntax and the semantics of object
% logics of interest, in which reasoning problems in concrete
% application domains can be modeled and solved with pen and paper. In
% fact, this approach can also be followed on the computer (using HOL as
% a formal meta-language) for even very challenging object logics (such
% as HOML) to enable interactive and automated theorem proving with
% existing theorem provers for HOL.

% Just as commonly the case in logic textbooks, in the embeddings
% approach classical higher-order logic is utilised as a meta-language
% to encode the syntax and the semantics of object logics in which the
% proof problems of interest are then modeled in. Modulo the embedding
% state of the art automated theorem provers for classical higher-order
% logic can then be utilised as object logic reasoning tools.

In our experiments in this branch of metaphysics we
utilise an embedding of HOMLs, such as \textbf{K}, \textbf{KB} and
\textbf{S5} with various domain conditions (possibilist and actualist
quantification), in HOL. More precisely, formulas in HOML are \emph{lifted}, i.e., converted
into predicates over worlds, which are themselves explicitly
represented as terms. The logical constants of HOML are translated to
HOL terms in such a way that, for instance,
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation,
%\cite{DBLP:journals/logcom/Ohlbach91}, 
which is here intra-logically realized --- and extended for
quantifiers --- in HOL by stating a set of equations defining the
logical constants. The resulting object logic is
the HOML \textbf{K} with rigid terms and constant domains (possibilist
quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
using an existence predicate to guard the quantifiers. 
%The embedding
%approach is, therefore, very flexible.

% \subsection{Improved Embedding}\label{sec:improvedembedding}
% \begin{figure}[t]
% \centerline{\includegraphics[width=1\columnwidth,height=10cm]{./QML_S5U.png}}
% \caption{Improved Embedding of \SFiveU} \label{QML_S5U}
% \end{figure}

% The modal logic \SFive requires that the accessibility relation be
% reflexive, symmetric and transitive. The usual approach to embed
% \SFive would be to use the standard translation for \textbf{K} described above
% and to state that $r$ is an equivalence relation, e.g., by postulating
% the following axioms:
% \begin{itemize}
% \item Reflexivity: $\forall x. (r~x~x)$
% \item Symmetry: $\forall x. \forall y. (r~x~y) \rightarrow (r~y~x)$ 
% \item Transitivity: $\forall x. \forall y. \forall z. (r~x~y) \wedge (r~y~z) \rightarrow (r~x~z)$
% \end{itemize}
% Instead, we consider here an alternative description, that we call \SFiveU, based on the following condition on $r$:
% \begin{itemize}
% \item Universality: $\forall x. \forall y. (r~x~y)$
% \end{itemize}

% It is important to note that $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi$ 
% \cite{Blackburn}, and therefore \SFive and \SFiveU are traditionally considered to be 
% two different descriptions of the same modal logic. Nevertheless, \SFive and \SFiveU 
% differ in the shapes of frames they admit:
% %It is easy to see that \SFiveU is at least as strong as \SFive:
% %universality entails that $r$ is an equivalence relation. \SFiveU is,
% %in fact, \emph{strictly stronger} than \SFive. 
% \SFiveU only admits
% complete\footnote{A graph is \emph{complete} iff there is a directed
%   edge connecting every ordered pair of vertices.} frames, whereas
% \SFive admits non-complete frames as long as all their components are
% complete. In other words, in \SFiveU we face one single equivalence
% class of possible worlds, while in \SFive we may face several
% disconnected equivalence classes. In fact, for this reason, \SFiveU is
% considered as metaphysically more appropriate by some philosophers;
% cf. \cite[p.~127]{williamson13}.

% Furthermore, for \SFiveU an improved embedding is possible. Universality implies
% that the guarding predicates in the definitions of $\Box$ and
% $\Diamond$ always hold. Therefore, they can be omitted and the
% accessibility relation can be dispensed altogether. The modal
% operators can then be defined merely as:
% % \begin{itemize}
% % \item $\BlackBox \varphi \equiv \forall w. (\varphi w)$ 
% % \item $\BlackDia \varphi \equiv \exists w. (\varphi w)$
% % \end{itemize}
% \[\Box \varphi \equiv \lambda w. \forall v.  \varphi(v) \qquad \text{and}
% \qquad \Dia \varphi \equiv \lambda w.\exists v. \varphi(v)\]

% The new embedding of \SFiveU in Isabelle/HOL is shown in
% Fig.~\ref{QML_S5U}.
% With this improved embedding, the final theorem T3
% (\textit{Necessarily, there exists God}) can be derived from
% Scott's consistent version of the axioms fully automatically.  The fully
% automatic proof has been generated (in about 2.5 seconds) by the theorem prover
% \textsc{Leo-II}~\cite{leo2} and subsequently verified in the proof
% assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5U}.
% The collaboration between the two systems has been orchestrated by
% Isabelle's Sledgehammer tool \cite{Sledgehammer}.

% With the embedding used by Benzm\"uller and Woltzenlogel-Paleo \cite{C40}, the
% provers still had to be given the intermediate theorem T2 and the corollary
% C in order to manage to prove T3.

% Another evidence that the new embedding provides a significant performance boost 
% is the successful automatic verification in Isabelle/HOL (with its automatic tactic Meson) of the modal
% collapse \cite{Sobel}, which is one of the most strongly criticized
% `side-effects' of G\"odel's and Scott's variants of the proof. In previous work \cite{C40}
% the modal collapse has been proven by the higher-order provers
% \textsc{Satallax} \cite{Satallax} and \textsc{Leo-II}, but a fully automatic
% verification in the highly trusted Isabelle/HOL system still failed
% \cite{J28}.  The success with the new embedding can be seen in Fig.~\ref{Scott_S5U}.

% \subsection{Improved Syntax in Isabelle}\label{sec:improvedsyntax}

% Wider adoption of HOL theorem proving technology for
% reasoning about and within embedded object logics, especially among
% non-expert users, is still hindered by the gap between the syntax used
% by people, when they write logical formulas with pen and paper, and
% the syntax used by HOL theorem provers. Even when the syntax of the
% underlying higher-order system is elegant (as is the case in
% Isabelle/HOL), the embedding of HOML into HOL may easily expose
% details of HOL that may be uncommon to the user, disturbing his/her
% experience while using the system. To illustrate this point,
% Fig.~\ref{UglyEssence} shows how the definition of essence looked like
% in previous work \cite{J28}, where advanced syntax-sugaring features
% were not used. It looks notably higher-order, and its style differs
% significantly from the common style seen in works on modal logics and
% the ontological argument. The following specific issues can be
% enumerated:
% \begin{enumerate}
% \item $\lambda$-abstractions, which are typically a HOL feature, appear explicitly in places where they did not need to in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}).
% \item Quantifiers appear as higher-order defined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
% \item The lifted modal connectives are represented by prefixing the
%   letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
%   user, as it constantly reminds him/her that there is something
%   unusual about the modal connectives.
% \item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
% Instead of reading $\psi(y)$, as he/she would expect, he/she has to read $(\psi ~ y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
% \end{enumerate}

% \begin{figure}
% \centerline{\includegraphics[width=1\columnwidth]{./UglyEssence.png}}
% \caption{Definition of Essence using Old Syntax} \label{UglyEssence}
% \end{figure}

% In the embedding presented here, in Fig.~\ref{QML_S5U}, advanced
% syntax-sugaring effects provided by Isabelle were used to prevent
% issues as those enumerated above. The possibility to define boldface connectives allows us to drop the prefix; ``\texttt{binder}'' annotations enable modal quantifiers to be used in the standard binding way and reduce the need for explicit lambda abstractions; and a careful choice of priorities for infix connectives gives the parenthesis conventions that are more familiar to the user. As desired, the definition of essence in Fig.~\ref{Scott_S5U} is undeniably more immediately recognizable and comprehensible than the definition in Fig.~\ref{UglyEssence}. The embedding technique is now completely transparent to the user.

% The syntax improvements described here render the computer-assisted analysis of ontological arguments accessible to a wider audience and ease the adoption of logic-based artificial intelligence tools by philosophers interested in topics where modal logic reasoning is required.

@@

\pagebreak

\paragraph{Argument Reconstruction in Isabelle.}  \label{sec:arg2}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[t]{0.715\textwidth}
    \includegraphics[width=\textwidth]{./Manuscript2.png}
    \caption{G\"{o}del's manuscript, with mutually inconsistent axioms and
      definitions highlighted (with permission from the Kurt G\"odel Papers, Shelby White and Leon Levy Archives Center, Princeton, NJ, USA, on deposit at Princeton University)} \label{GoedelScript} 
  \end{subfigure}
   \begin{subfigure}[t]{0.28\textwidth}
     \includegraphics[width=\textwidth,height=7.2cm]{./Inconsistency_S5U_direct.png}
     \caption{Inconsistency in HOML \SFiveU} \label{Inconsistency_S5U} 
   \end{subfigure}
% \includegraphics[width=.73\textwidth]{./Images/Manuscript2.png} \hfill
% \includegraphics[width=.26\textwidth]{./Images/Inconsistency_S5U_direct.png}
 \caption{The inconsistency in G\"{o}del's manuscript has been
   detected and verified by HOL ATPs} 
\end{figure*}
To verify the correctness of the informal argument explained above, it
was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
  external provers such as \textsc{Leo-II} or Satallax, 
  constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
inessential parts (cf. Fig.~\ref {InconsistencyIsabelleK}). The essential use of the Empty Essence Lemma, on
the other hand, is explicitly stated, to ensure that Isabelle is
reconstructing the same argument. In fact, without the help of this
lemma, Metis is still not strong enough to refute G\"odel's
axioms.

\paragraph{Mapping the Inconsistency to G\"odel.}
The inconsistency verified in Fig.~\ref{InconsistencyIsabelleK} follows from the definition of
essence (ess), the definition of necessary existence (NE), the
axioms A1a and A2 (which entail theorem T1), and axiom A5. It remains to show that
these ingredients are actually present in G\"odel's manuscript in
Fig.~\ref{GoedelScript}.

This can be easily seen: Axiom A1a in
Fig.~\ref{InconsistencyIsabelleK} is implied by Axiom Ax2 and the
highlighted footnote remark in Fig.~\ref{GoedelScript}. Axioms A2 and
A5 in Fig.~\ref{InconsistencyIsabelleK} correspond to Ax4 and Ax3 in
Fig.~\ref{GoedelScript}. The definitions of essence and necessary
existence are easy to identify. Therefore, the verified
inconsistency from Fig.~\ref{InconsistencyIsabelleK} does apply to 
G\"odel's original manuscript.

\paragraph{Inconsistency of G\"odel's Axioms in \SFiveU.}
Isabelle/HOL's Sledgehammer tool, which orchestrates calls to
external provers such as \textsc{Leo-II}, still
fails to detect the inconsistency of G\"odel's axioms in the standard
embedding of \SFive, while a direct modeling of the problem in TPTP THF syntax
in combination with a direct call of \textsc{Leo-II} succeeded. In
other words, without independent experiments with no mediation through Sledgehammer, the
inconsistency would not have been detected.

%\marginpar{ToDo: Are we really supposed to use \S instead of ``Sec.''?}
On the other hand (and further confirming the claims from \S\ref{sec:improvedembedding}), 
the reconstruction in Isabelle/HOL with the improved embedding for \SFiveU was more efficient: 
the inconsistency could be detected
by \textsc{Leo-II} also when called via
Sledgehammer. Moreover, the result could subsequently be verified with
Metis even without the Empty Essence Lemma (cf. Fig.~\ref{Inconsistency_S5U}). 
% Interestingly, Metis' argument is different from that of
% \S\ref{sec:arg1} and \S\ref{sec:arg2}, since axiom A1b is not used.

>>>>

@@

In our experiments in this branch of metaphysics we
utilise an embedding of HOMLs, such as \textbf{K}, \textbf{KB} and
\textbf{S5} with various domain conditions (possibilist and actualist
quantification), in HOL. More precisely, formulas in HOML are \emph{lifted}, i.e., converted
into predicates over worlds, which are themselves explicitly
represented as terms. The logical constants of HOML are translated to
HOL terms in such a way that, for instance,
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation,
%\cite{DBLP:journals/logcom/Ohlbach91}, 
which is here intra-logically realized --- and extended for
quantifiers --- in HOL by stating a set of equations defining the
logical constants. The resulting object logic is
the HOML \textbf{K} with rigid terms and constant domains (possibilist
quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
using an existence predicate to guard the quantifiers.

@@

% \pagebreak

% \paragraph{Argument Reconstruction in Isabelle.}  \label{sec:arg2}
% \begin{figure*}[t]
%   \centering
%   \begin{subfigure}[t]{0.715\textwidth}
%     \includegraphics[width=\textwidth]{./Manuscript2.png}
%     \caption{G\"{o}del's manuscript, with mutually inconsistent axioms and
%       definitions highlighted (with permission from the Kurt G\"odel Papers, Shelby White and Leon Levy Archives Center, Princeton, NJ, USA, on deposit at Princeton University)} \label{GoedelScript} 
%   \end{subfigure}
%    \begin{subfigure}[t]{0.28\textwidth}
%      \includegraphics[width=\textwidth,height=7.2cm]{./Inconsistency_S5U_direct.png}
%      \caption{Inconsistency in HOML \SFiveU} \label{Inconsistency_S5U} 
%    \end{subfigure}
% % \includegraphics[width=.73\textwidth]{./Images/Manuscript2.png} \hfill
% % \includegraphics[width=.26\textwidth]{./Images/Inconsistency_S5U_direct.png}
%  \caption{The inconsistency in G\"{o}del's manuscript has been
%    detected and verified by HOL ATPs} 
% \end{figure*}
% To verify the correctness of the informal argument explained above, it
% was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
%   external provers such as \textsc{Leo-II} or Satallax, 
%   constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
% inessential parts (cf. Fig.~\ref {InconsistencyIsabelleK}). The essential use of the Empty Essence Lemma, on
% the other hand, is explicitly stated, to ensure that Isabelle is
% reconstructing the same argument. In fact, without the help of this
% lemma, Metis is still not strong enough to refute G\"odel's
% axioms.

% \paragraph{Mapping the Inconsistency to G\"odel.}
% The inconsistency verified in Fig.~\ref{InconsistencyIsabelleK} follows from the definition of
% essence (ess), the definition of necessary existence (NE), the
% axioms A1a and A2 (which entail theorem T1), and axiom A5. It remains to show that
% these ingredients are actually present in G\"odel's manuscript in
% Fig.~\ref{GoedelScript}.

% This can be easily seen: Axiom A1a in
% Fig.~\ref{InconsistencyIsabelleK} is implied by Axiom Ax2 and the
% highlighted footnote remark in Fig.~\ref{GoedelScript}. Axioms A2 and
% A5 in Fig.~\ref{InconsistencyIsabelleK} correspond to Ax4 and Ax3 in
% Fig.~\ref{GoedelScript}. The definitions of essence and necessary
% existence are easy to identify. Therefore, the verified
% inconsistency from Fig.~\ref{InconsistencyIsabelleK} does apply to 
% G\"odel's original manuscript.

% \paragraph{Inconsistency of G\"odel's Axioms in \SFiveU.}
% Isabelle/HOL's Sledgehammer tool, which orchestrates calls to
% external provers such as \textsc{Leo-II}, still
% fails to detect the inconsistency of G\"odel's axioms in the standard
% embedding of \SFive, while a direct modeling of the problem in TPTP THF syntax
% in combination with a direct call of \textsc{Leo-II} succeeded. In
% other words, without independent experiments with no mediation through Sledgehammer, the
% inconsistency would not have been detected.

% %\marginpar{ToDo: Are we really supposed to use \S instead of ``Sec.''?}
% On the other hand (and further confirming the claims from \S\ref{sec:improvedembedding}), 
% the reconstruction in Isabelle/HOL with the improved embedding for \SFiveU was more efficient: 
% the inconsistency could be detected
% by \textsc{Leo-II} also when called via
% Sledgehammer. Moreover, the result could subsequently be verified with
% Metis even without the Empty Essence Lemma (cf. Fig.~\ref{Inconsistency_S5U}). 
% % Interestingly, Metis' argument is different from that of
% % \S\ref{sec:arg1} and \S\ref{sec:arg2}, since axiom A1b is not used.
